An update and reassessment of fern and lycophyte diversity data in the Japanese Archipelago

### 2019-06-25 manuscript/japan_ferns_diversity_ms.Rmd
Add headers, output to word

---
title: "An update and reassessment of fern and lycophyte diversity data in the Japanese Archipelago"
output: html_document
---

@@

Introduction
Ferns and lycophytes or seed-free vascular plants, also traditionally known as ‘pteridophytes’ are a group of plants which have been paid strong attention as research material of biology. Although estimated number of extant species of ferns and lycophytes is only 11916 (Pteridophyte Phylogeny Group, 2016), much less (approximately one twelfth) than seed plant diversity, the frequency of publication of studies used ferns and lycophytes especially in the field of systematics and evolution seems to be high comparing their proportion on number of species. Thus, ferns and lycophytes are relatively rich organisms for information on their diversity. In last centenary, the diversity of ferns and lycophytes in Japan was quite thoroughly studied, and is probably the third best studied area after Europe and North America. Guo et al. (2003) already noticed that the ferns and lycophyte flora in Japan accompanied with rich trait information was worth analyzing determinants of species diversity, abundance and distribution. Even today, there are scarce examples of analyses on fern diversity pattern in a country or wider scale (e.g. Bogonovich et al. 2014, Mountier et al. 2018). A sole serious problem in the analyses by Gao et al. (2003) was the lack of phylogenetic information at that time, but it was completely solved by subsequent studies (Ebihara 2011, Ebihara et al. 2010). In addition, we should note that the occurrence data they used based on Kurata and Nakaike (1979, 1981, 1983, 1985, 1987, 1990, 1994, 1997) did not covered all the species. Reflecting numerous new occurrences and recircumscription of the species based on the results of recent studies, Ebihara (2016, 2017) provided updated distribution maps for all the native taxa in Japan. We intend to publicize up-to-date diversity datasets on ferns and lycophytes in Japan combined with a review of recent progresses and reassessment of the results in Guo et al. (2003).

Flora
 The studies on fern and lycophyte flora of the Japanese Archipelago initiated with 44 species enumerated in Thunberg (1786)’s “Flora Japonica”. Thereafter the number of recognized species native to there are gradually increasing: 460 species including subspecies and varieties (Makino and Nemoto 1925), ca. 400 species excluding those in Ogasawara (Bonin) and Ryukyu Islands (Tagawa 1959) and 630 species (Iwatsuki 1995). Our latest checklist of native species of Japan ‘FernGreenList ver. 1.01’ (Ebihara et al. 2017a) comprises 691 species (721 taxa including subspecies and varieties [Electronic supplementary material 1]; hereafter call ‘taxa’) belonging to 37 of the 51 families accepted in PPGI (Pteridophyte Phylogeny Group 2016). As almost no uninvestigated area remains in the archipelago, most of the new additions to the flora after the 1990’s are newly segregated species from known ones based on evidences from ploidy levels, enzyme analyses and/or DNA sequences (e.g. Ebihara et al. 2019a, Fujiwara et al. 2018, Hori et al. 2015, 2018a, 2018c, Lin et al. 1994, 1996, Masuyama and Watano 2010, Murakami et al. 1999, Ohta and Takamiya 1999, Serizawa 2015, Takamiya et al. 1997) along with only small number of brand-new records (Ebihara et al. 2014, 2019a). The number of taxa in Japan is nearly identical to that in Taiwan (721 taxa excluding forms and hybrids, Knapp 2014). The strongest floristic affinity is found to the continental part of China sharing 487 taxa (67.5%), almost corresponding to two thirds of the native taxa in Japan (Ebihara 2016, 2017). Accumulated up-to-date information identified 125 endemic taxa (17.3% of the native taxa) to Japan (Ebihara 2018).

Interspecific hybrids
 Hybrid ferns have been widely utilized as material of evolutionary biological studies (Sigel 2016), and recognition of interspecific hybrids is always their starting points. Besides those 721 taxa, a huge number of interspecific natural hybrid taxa (371 combinations are listed in Ebihara 2017) is an outstanding characteristic of the flora. This is much more than known hybrid combinations in North America (xxx), China (10 were formerly accepted in Flora of China [Wu et al. 2013] and 20 or more combinations were suggested in the notes) and Taiwan (13 listed in Knapp 2014). Such high diversity is most likely a result of efforts on seeking new hybrid combinations by Satoru Kurata (1922-1978) and his pupils, but might be, at least in part, by a geographical effect of the long archipelago extending north to south which enable sympatric occurrence of species with different habitat preferences according climate fluctuation (any example???). It is highly likely that most of the interspecific hybrid individuals in a large part of the world are unrecognized yet, including the cases of miscategorization to non-hybrid taxa. We should mind that only a few combinations of hybridity were verified by artificial crossing, enzyme patterns or nuclear DNA markers and the others are still putative hybrids on morphological basis. An expedient method observing irregular size and shape of spores is often adopted for recognizing interspecific F1 hybrids, but is not always true represented by normal-shaped regular spores produced by homoploid hybrids, Botrychium x silvicola (Ebihara et al. 2015, Sahashi 1983) and other putative homoploid hybrids of Botrychium species in Japan (N. Sahashi, personal communication). An incomplete establishment of reproductive isolation was observed between Osmunda lancea, a rheophytic species endemic to Japan and its closest relative O. japonica, and their ‘hybrid’ O. x intermedia is actually hybrid swarms including not only F1 but also F2 or later-generation hybrid (Yatabe et al. 2009, 2011). Fern interspecific hybrids have long been believed to occur in places where sporophytes of their two parent species are sympatric, and this idea was an important ground for identification of putative parentages. However, our studies demonstrated a number of examples of sterile interspecific hybrid unaccompanied by their parental species. One of the extreme cases is found in a dennstaedtioid fern Monachosorum x arakii which was originally considered as an endemic species to Japan. Our study (Ebihara et al. 2016) concluded that it is a sterile hybrid between Monachosorum henryi and M. nipponica, the former is not currently distributed in Japan but in Taiwan, continental China and Himalaya. Although it is somewhat a special case with strong vegetative reproduction ability not only with rhizomes but also with bulbils making relict distribution possible, we should always hypothesize any possible combination of hybrid parentages as flexible as possible. Apart from relict distribution, we obtained strong evidences for contribution of independent gametophytes to formation of hybrid sporophytes in Vandenboschia (Ebihara et al. 2009).

Occurrence data and regional species richness
Eight volumes of "Illustrations of Pteridophytes of Japan" (Kurata and Nakaike, 1979, 1981, 1983, 1985, 1987, 1990, 1994, 1997) were a pioneer work of a nationwide organism distribution survey by citizens in Japan. Along with brief description about each species, these books featured 10 km-grid based distribution maps vouchered by ca. 180000 specimens collected by the members of the Nippon Fernist Club. A large part of the specimens presently deposited in the herbarium of National Museum of Nature and Science (TNS) were recently re-identified accommodating up-to-date species circumscription, and ca. 35000 specimens which are new records on the 10 km-grids were newly collected by the club members. Occurrences based on specimens are usually regarded as presence-only data, but present grid maps can be virtually regarded as presence/absence data as the results of the members’ extensive searching in the field. For the convenience in reusing the occurrence data, a list of specimens which assure the distribution maps in Ebihara (2016, 2017) are publicized here, and it contains 216687 herbarium specimen records (Electronic supplementary material 2). We should note that our policy for the dataset is to include specimens or specimen-images examined by ourselves (only exception is Isoetes occurrences which cited from a specimens list provided by Takamiya et al. [1997]). At least one species is recorded in 4343 of 4853 grids (89.5 %), and the species occurred in the maximum number of grids is Pteridium aquilinum subsp. japonicum (Dennstaedtiaceae) recorded in 2777 grids (57.2%).
Even not included in present dataset, much larger number of specimen data without images are available on-line via Science Museum Net (http://science-net.kahaku.go.jp/) which is a portal of natural history specimens deposited in museums in Japan, and the occurrence data are also searchable in the GBIF data portal (https://www.gbif.org/). As of May 23, 2019, Japan is the country of locality with the largest number of fern and lycophyte specimens in GBIF, totally 531311 occurrences.
In the summary of species richness in 10 km-grids excluding nothotaxa (Fig. 1a), the grids with the highest (212 taxa, grid code: 453034) and the second highest diversity (210 taxa, grid code: 453044) are found in the mountain area of Yakushima Island, located south of Kyushu including the island’s highest peak (Mt. Miyanoura, 1936 m in elevation). The third highest grid (202 taxa, grid code 503547) is located in the area covering Nachi-no-taki Falls in southern part of Wakayama Prefecture, southern Honshu.

@@

Cytotaxonomic data
Even today chromosome numbers of ferns and lycophytes which are important clues for knowing the plant’s ploidy level and reproductive mode as well as basic chromosome numbers, which were used as information to infer phylogenetic affinities before the introduction of DNA sequencing. Researchers in Japan stimulated by Irene Manton’s leading work “Problems of Cytology and Evolution in the Pteridophyta” (Manton 1950) actively made chromosome counts of ferns and lycophytes using the squash technique in the second half of the 20th century, and more than 150 papers had been published – all the counts on Japanese material before 1996 was indexed by Takamiya (1996). Additions have been continuously made, and Nakato and Ebihara (2016) summarized that the information drawn from Japanese individuals is available for 74 % of the native taxa (61 %, if nothotaxa are counted) – it is a much higher coverage compared with other areas of East Asia (e.g. ca. 27 % in China, Cheng and Zhang [2010]) and the second highest coverage in Asia following India (ca. 89 %, Bir and Verma [2010]). Further addition is highly needed for ascertaining some early records without preserving voucher specimen, and for detecting infraspecific polyploidy and/or multiple reproductive modes overlooked by previous superficial samplings as well as unsampled taxa. Flow cytometry is recently used for inferring ploidy levels of large number of samples belonging to a closely related species groups (e.g. Ebihara et al. 2005, Fujiwara et al. 2018, Hori et al. 2015, Nitta et al. 2011), but the method always requires at least one control sample whose ploidy level was confirmed by chromosome counts. Flow cytometry analyses using not only fresh leaves but also recently collected spores in herbarium specimens, succeeded in some ferns (Kuo and Huang 2017), might accelerate data accumulation.

Threatened species
 Latest national red list of Japan (Ministry of the Environment, Japan 2019) included more than one third of the native fern and lycophyte taxa (260 in total, 6 EX, 3 EW, 85 CR, 59 EN, 68 VU, 38 NT and 1 DD; see Fig. 1f for the distribution of richness). The largest threat for ferns is herbivory of the sika deer distributed throughout the archipelago which obviously increased after 1980’s (Takatsuki 2009). In particular, species growing on the forest floor (e.g. Athyrium, Terada and Takamiya 2006) undergo severe damage, and several narrowly distributed species in Kyushu and Yakushima Island became almost extinct or only found inside deer fences. Although national law for endangered species is recently expanding objective fern and lycophyte species which are proscribed collection and sellout (28 species as of May 2019), it is apparently not directly effective for preventing herbivory of deer. There are still a few reports of conservation activities and researches on ferns and lycophytes in Japan: ubiquitous genotyping of Athyrium viridescentipes (Izuno et al. 2012) and clarifying origin and propagation from spore in Dryopteris shibipedis (Ebihara et al. 2012) which was extinct in the wild. Despite 70 of the 260 taxa in the latest national red list are endemic species to Japan, they are neither assessed nor listed in the IUCN Red List as globally threatened species as of May 2019 (IUCN 2019).

Introduced species
Compared with flowering plants, number of naturalized species of ferns and lycophytes is quite small in the Japanese Archipelago, and they are almost uninvolved with the invasive species issue. Selaginella uncinata (Selaginellaceae) and Pityrogramma calomelanos (Pteridaceae) are two of scarce examples of relatively widespread introduced species in the country (Ebihara 2016). Dryopteris intermedia (ref), D. carthusiana (ref) (Dryopteridaceae) and Dennstaedtia punctilobula (Dennstaedtiaceae) (Ueno et al. 2019) were probably introduced with sprayed seeds on banks and only collected in each one or a few localities. Azolla cristata (Salviniaceae) is the only invasive fern species controlled by national low (Invasive Alien Species Act), but is often troubled in identification due to morphologically similarlity to an endangered and endemic species, A. japonica. Another similar species, Azolla filiculoides is still controversial on its natural distribution in Japan (Ebihara 2016, Suzuki et al. 2005). Selaginella moellendolffii (Selaginellaceae), Psilotum nudum (Psilotaceae), Adiantum capillis-veneris, Pteris vittata (Pteridaceae) and Thelypteris dentata (Thelypteridaceae) have certain natural distribution in the subtropical areas of the country, but are more frequently found as escaped plants from greenhouses. Murakami et al. (2008) demonstrated a range expansion of Thelypteris dentata facilitated by urbanization in Kinki District, Japan. The natural habitat of Cyrtomium falcatum subsp. falcatum is restricted to coastal areas (Ebihara et al. 2017a, Matsumoto 2003), and scattered occurrences in urban environments (Ebihara 2017) in inlands are probably introduced ones.

Reticulate evolution
 Species complexes or aggregate species exhibiting wide ranges of continuous morphological variation are quite commonly known in ferns and lycophytes (Haufler 1996), and this trend seems to be true for the flora of Japan. In case of that interspecific F1 hybrids make the species boundaries ambiguous, we could usually find distinct gaps between them by removing their hybrid individuals suggested by irregular-shaped spores with low germination rates (e.g. Ohta and Takamiya 1999 in the Diplazium mettenianum complex). When interspecific hybrids overcome sterility by chromosome doubling, the situation becomes more complicated and the biological units composing the complex are not easily clarifiable solely by morphological approach. The allopolyploids of hybrid origins often re-hybridize with their parental species --- such "reticulate evolution" process became well-known after the study on Appalachain Asplenium (Wagner 1954). Apparently similar approach combining chromosome and morphological observations was rarely adopted to Japanese species. After 1990's, allozyme analyses were frequently applied to the fern complexes in Japan, and plastid sequences often accompanied as indicators of maternal lineage (e.g. Darnaedi et al. 1990 in Dryopteris yakusilvicola and its related species, Lin et al. 1996 in Odontosoria, Shinohara et al. 2010 in Lepisorus, Takamiya and Ohta 2001 in Diplazium). In 2000's, single or low copy nuclear DNA regions became popularly used as biparently inherited markers (e.g. Ishikawa et al. 2002), and sequence polymorphisms were efficiently isolated by cloning or the SSCP (Single Strand Conformation Polymorphism) method (initiated by Ebihara et al. 2005 in the Vandenboschia radicans complex, thereafter applied to various fern groups including Ceratopteris [Adjie et al. 2007]), the Pteris cretica complex [Jaruwattanaphan et al. 2013], the Davallia repens complex [Chen et al. 2014], the Asplenium normale complex [Fujiwara et al. 2017]). Although next-generation sequencing is widely used in phylogenetic and population genetic studies on various organisms in recent years, it is still not popularly applied to analyses on species complexes of ferns and lycophytes mainly due to technical difficulties in removing chimeral contigs identifying and homologous copies from the obtained sequences. More recently Rothfels et al. (2017) succeeded in phylogenetic analysis on a polyploid complex of cystoptreroid ferns using PacBio sequences, and the method is highly expected to apply wide ranges of fern complexes.

Species complexes involved with apomixis
 Further complexity is provided when apomictic reproduction (reviewed in Grusz et al. 2016) is concerned. The topic, apomixis or apogamy in ferns has attracted researchers in Japan as the reproductive mode is known in a higher rate in the native taxa of the country (13 % of the species whose reproductive modes are known in Takamiya 1996; 12 % in present study) than the average rate in the world (ca. 10 %, Grusz et al. 2016). Apomixis plays a role not only in overcoming sterility of interspecific hybrids (e.g. Park and Kato 2003) but also making secondary hybridization by fertilization between a sperm produced by the ‘hybrid’ and an egg of sexually reproducing races. This process can generates countless apomict races with various genomic combinations which are morphologically continuous, and the races are often controversial about circumscription of ‘species’ (a polymorphic species versus numerous morphologically ill-defined species). One of the best known apomict complexes troublesome in identification, the Dryopteris varia complex was recently clarified their internal relationships using single-copy nuclear DNA markers (Hori et al. 2014, 2015, 2016), and analyses on several other complexes are in progress. However, even using molecular markers, we often meet an obstacle that we cannot find progenitors or sexual diploids of the complex. Hori et al. (2018d) discovered several sexual diploid individuals of Dryopteris in southern China, and as the progenitors the newly found diploids well explained the reticulate evolution processes of the D. erythrosora complex in Japan which had bothered researchers for many years (Hori et al. 2018b). This case supported the hypothesis by Gao et al (2003): “The ancestral sexual species of many Japanese apogamous species occur in China, suggesting that these species might have migrated from China”. In contrast to these successful clarifications, we are still struggling with more challenging complexes. The species of Cyrtomium (Dryopteridaceae) in Japan typically match this pattern: 12 native species and subspecies are all triploid apomicts except for two subspecies of C. falcatum (subsp. australe and subsp. littorale, sexual diploid) and C. devexiscapulae (sexual tetraploid) (Ebihara et al. 2017a, Matsumoto 2003). Despite careful exploration of sexually reproducing individuals by examining spore number per sporangium in herbarium specimens collected in China, a large part of them are apomictic and sexual ones were found less frequently in geographically narrow ranges (Mitsuta 1986, 1988). The Pteris fauriei complex is probably in similar situation, and it includes several endemic ‘species’ to Japan which are diploid or triploid apomict (Ebihara 2016). The “Diploids-First Approach” (Beck et al. 2010) is a principle for the studies on fern reticulate complexes, but we should mind that diploids are not always present nearby including possible extinction in complexes involved with apomixis.

>>>>

---
title: "An update and reassessment of fern and lycophyte diversity data in the Japanese Archipelago"
output: word_document
---

@@

## Introduction

Ferns and lycophytes or seed-free vascular plants, also traditionally known as ‘pteridophytes’ are a group of plants which have been paid strong attention as research material of biology. Although estimated number of extant species of ferns and lycophytes is only 11916 (Pteridophyte Phylogeny Group, 2016), much less (approximately one twelfth) than seed plant diversity, the frequency of publication of studies used ferns and lycophytes especially in the field of systematics and evolution seems to be high comparing their proportion on number of species. Thus, ferns and lycophytes are relatively rich organisms for information on their diversity. In last centenary, the diversity of ferns and lycophytes in Japan was quite thoroughly studied, and is probably the third best studied area after Europe and North America. Guo et al. (2003) already noticed that the ferns and lycophyte flora in Japan accompanied with rich trait information was worth analyzing determinants of species diversity, abundance and distribution. Even today, there are scarce examples of analyses on fern diversity pattern in a country or wider scale (e.g. Bogonovich et al. 2014, Mountier et al. 2018). A sole serious problem in the analyses by Gao et al. (2003) was the lack of phylogenetic information at that time, but it was completely solved by subsequent studies (Ebihara 2011, Ebihara et al. 2010). In addition, we should note that the occurrence data they used based on Kurata and Nakaike (1979, 1981, 1983, 1985, 1987, 1990, 1994, 1997) did not covered all the species. Reflecting numerous new occurrences and recircumscription of the species based on the results of recent studies, Ebihara (2016, 2017) provided updated distribution maps for all the native taxa in Japan. We intend to publicize up-to-date diversity datasets on ferns and lycophytes in Japan combined with a review of recent progresses and reassessment of the results in Guo et al. (2003).

## Flora

The studies on fern and lycophyte flora of the Japanese Archipelago initiated with 44 species enumerated in Thunberg (1786)’s “Flora Japonica”. Thereafter the number of recognized species native to there are gradually increasing: 460 species including subspecies and varieties (Makino and Nemoto 1925), ca. 400 species excluding those in Ogasawara (Bonin) and Ryukyu Islands (Tagawa 1959) and 630 species (Iwatsuki 1995). Our latest checklist of native species of Japan ‘FernGreenList ver. 1.01’ (Ebihara et al. 2017a) comprises 691 species (721 taxa including subspecies and varieties [Electronic supplementary material 1]; hereafter call ‘taxa’) belonging to 37 of the 51 families accepted in PPGI (Pteridophyte Phylogeny Group 2016). As almost no uninvestigated area remains in the archipelago, most of the new additions to the flora after the 1990’s are newly segregated species from known ones based on evidences from ploidy levels, enzyme analyses and/or DNA sequences (e.g. Ebihara et al. 2019a, Fujiwara et al. 2018, Hori et al. 2015, 2018a, 2018c, Lin et al. 1994, 1996, Masuyama and Watano 2010, Murakami et al. 1999, Ohta and Takamiya 1999, Serizawa 2015, Takamiya et al. 1997) along with only small number of brand-new records (Ebihara et al. 2014, 2019a). The number of taxa in Japan is nearly identical to that in Taiwan (721 taxa excluding forms and hybrids, Knapp 2014). The strongest floristic affinity is found to the continental part of China sharing 487 taxa (67.5%), almost corresponding to two thirds of the native taxa in Japan (Ebihara 2016, 2017). Accumulated up-to-date information identified 125 endemic taxa (17.3% of the native taxa) to Japan (Ebihara 2018).

## Interspecific hybrids

Hybrid ferns have been widely utilized as material of evolutionary biological studies (Sigel 2016), and recognition of interspecific hybrids is always their starting points. Besides those 721 taxa, a huge number of interspecific natural hybrid taxa (371 combinations are listed in Ebihara 2017) is an outstanding characteristic of the flora. This is much more than known hybrid combinations in North America (xxx), China (10 were formerly accepted in Flora of China [Wu et al. 2013] and 20 or more combinations were suggested in the notes) and Taiwan (13 listed in Knapp 2014). Such high diversity is most likely a result of efforts on seeking new hybrid combinations by Satoru Kurata (1922-1978) and his pupils, but might be, at least in part, by a geographical effect of the long archipelago extending north to south which enable sympatric occurrence of species with different habitat preferences according climate fluctuation (any example???). It is highly likely that most of the interspecific hybrid individuals in a large part of the world are unrecognized yet, including the cases of miscategorization to non-hybrid taxa. We should mind that only a few combinations of hybridity were verified by artificial crossing, enzyme patterns or nuclear DNA markers and the others are still putative hybrids on morphological basis. An expedient method observing irregular size and shape of spores is often adopted for recognizing interspecific F1 hybrids, but is not always true represented by normal-shaped regular spores produced by homoploid hybrids, Botrychium x silvicola (Ebihara et al. 2015, Sahashi 1983) and other putative homoploid hybrids of Botrychium species in Japan (N. Sahashi, personal communication). An incomplete establishment of reproductive isolation was observed between Osmunda lancea, a rheophytic species endemic to Japan and its closest relative O. japonica, and their ‘hybrid’ O. x intermedia is actually hybrid swarms including not only F1 but also F2 or later-generation hybrid (Yatabe et al. 2009, 2011). Fern interspecific hybrids have long been believed to occur in places where sporophytes of their two parent species are sympatric, and this idea was an important ground for identification of putative parentages. However, our studies demonstrated a number of examples of sterile interspecific hybrid unaccompanied by their parental species. One of the extreme cases is found in a dennstaedtioid fern Monachosorum x arakii which was originally considered as an endemic species to Japan. Our study (Ebihara et al. 2016) concluded that it is a sterile hybrid between Monachosorum henryi and M. nipponica, the former is not currently distributed in Japan but in Taiwan, continental China and Himalaya. Although it is somewhat a special case with strong vegetative reproduction ability not only with rhizomes but also with bulbils making relict distribution possible, we should always hypothesize any possible combination of hybrid parentages as flexible as possible. Apart from relict distribution, we obtained strong evidences for contribution of independent gametophytes to formation of hybrid sporophytes in Vandenboschia (Ebihara et al. 2009).

## Occurrence data and regional species richness

Eight volumes of "Illustrations of Pteridophytes of Japan" (Kurata and Nakaike, 1979, 1981, 1983, 1985, 1987, 1990, 1994, 1997) were a pioneer work of a nationwide organism distribution survey by citizens in Japan. Along with brief description about each species, these books featured 10 km-grid based distribution maps vouchered by ca. 180000 specimens collected by the members of the Nippon Fernist Club. A large part of the specimens presently deposited in the herbarium of National Museum of Nature and Science (TNS) were recently re-identified accommodating up-to-date species circumscription, and ca. 35000 specimens which are new records on the 10 km-grids were newly collected by the club members. Occurrences based on specimens are usually regarded as presence-only data, but present grid maps can be virtually regarded as presence/absence data as the results of the members’ extensive searching in the field. For the convenience in reusing the occurrence data, a list of specimens which assure the distribution maps in Ebihara (2016, 2017) are publicized here, and it contains 216687 herbarium specimen records (Electronic supplementary material 2). We should note that our policy for the dataset is to include specimens or specimen-images examined by ourselves (only exception is Isoetes occurrences which cited from a specimens list provided by Takamiya et al. [1997]). At least one species is recorded in 4343 of 4853 grids (89.5 %), and the species occurred in the maximum number of grids is Pteridium aquilinum subsp. japonicum (Dennstaedtiaceae) recorded in 2777 grids (57.2%).
Even not included in present dataset, much larger number of specimen data without images are available on-line via Science Museum Net (http://science-net.kahaku.go.jp/) which is a portal of natural history specimens deposited in museums in Japan, and the occurrence data are also searchable in the GBIF data portal (https://www.gbif.org/). As of May 23, 2019, Japan is the country of locality with the largest number of fern and lycophyte specimens in GBIF, totally 531311 occurrences.
In the summary of species richness in 10 km-grids excluding nothotaxa (Fig. 1a), the grids with the highest (212 taxa, grid code: 453034) and the second highest diversity (210 taxa, grid code: 453044) are found in the mountain area of Yakushima Island, located south of Kyushu including the island’s highest peak (Mt. Miyanoura, 1936 m in elevation). The third highest grid (202 taxa, grid code 503547) is located in the area covering Nachi-no-taki Falls in southern part of Wakayama Prefecture, southern Honshu.

@@

## Cytotaxonomic data

Even today chromosome numbers of ferns and lycophytes which are important clues for knowing the plant’s ploidy level and reproductive mode as well as basic chromosome numbers, which were used as information to infer phylogenetic affinities before the introduction of DNA sequencing. Researchers in Japan stimulated by Irene Manton’s leading work “Problems of Cytology and Evolution in the Pteridophyta” (Manton 1950) actively made chromosome counts of ferns and lycophytes using the squash technique in the second half of the 20th century, and more than 150 papers had been published – all the counts on Japanese material before 1996 was indexed by Takamiya (1996). Additions have been continuously made, and Nakato and Ebihara (2016) summarized that the information drawn from Japanese individuals is available for 74 % of the native taxa (61 %, if nothotaxa are counted) – it is a much higher coverage compared with other areas of East Asia (e.g. ca. 27 % in China, Cheng and Zhang [2010]) and the second highest coverage in Asia following India (ca. 89 %, Bir and Verma [2010]). Further addition is highly needed for ascertaining some early records without preserving voucher specimen, and for detecting infraspecific polyploidy and/or multiple reproductive modes overlooked by previous superficial samplings as well as unsampled taxa. Flow cytometry is recently used for inferring ploidy levels of large number of samples belonging to a closely related species groups (e.g. Ebihara et al. 2005, Fujiwara et al. 2018, Hori et al. 2015, Nitta et al. 2011), but the method always requires at least one control sample whose ploidy level was confirmed by chromosome counts. Flow cytometry analyses using not only fresh leaves but also recently collected spores in herbarium specimens, succeeded in some ferns (Kuo and Huang 2017), might accelerate data accumulation.

## Threatened species

Latest national red list of Japan (Ministry of the Environment, Japan 2019) included more than one third of the native fern and lycophyte taxa (260 in total, 6 EX, 3 EW, 85 CR, 59 EN, 68 VU, 38 NT and 1 DD; see Fig. 1f for the distribution of richness). The largest threat for ferns is herbivory of the sika deer distributed throughout the archipelago which obviously increased after 1980’s (Takatsuki 2009). In particular, species growing on the forest floor (e.g. Athyrium, Terada and Takamiya 2006) undergo severe damage, and several narrowly distributed species in Kyushu and Yakushima Island became almost extinct or only found inside deer fences. Although national law for endangered species is recently expanding objective fern and lycophyte species which are proscribed collection and sellout (28 species as of May 2019), it is apparently not directly effective for preventing herbivory of deer. There are still a few reports of conservation activities and researches on ferns and lycophytes in Japan: ubiquitous genotyping of Athyrium viridescentipes (Izuno et al. 2012) and clarifying origin and propagation from spore in Dryopteris shibipedis (Ebihara et al. 2012) which was extinct in the wild. Despite 70 of the 260 taxa in the latest national red list are endemic species to Japan, they are neither assessed nor listed in the IUCN Red List as globally threatened species as of May 2019 (IUCN 2019).

## Introduced species

Compared with flowering plants, number of naturalized species of ferns and lycophytes is quite small in the Japanese Archipelago, and they are almost uninvolved with the invasive species issue. Selaginella uncinata (Selaginellaceae) and Pityrogramma calomelanos (Pteridaceae) are two of scarce examples of relatively widespread introduced species in the country (Ebihara 2016). Dryopteris intermedia (ref), D. carthusiana (ref) (Dryopteridaceae) and Dennstaedtia punctilobula (Dennstaedtiaceae) (Ueno et al. 2019) were probably introduced with sprayed seeds on banks and only collected in each one or a few localities. Azolla cristata (Salviniaceae) is the only invasive fern species controlled by national low (Invasive Alien Species Act), but is often troubled in identification due to morphologically similarlity to an endangered and endemic species, A. japonica. Another similar species, Azolla filiculoides is still controversial on its natural distribution in Japan (Ebihara 2016, Suzuki et al. 2005). Selaginella moellendolffii (Selaginellaceae), Psilotum nudum (Psilotaceae), Adiantum capillis-veneris, Pteris vittata (Pteridaceae) and Thelypteris dentata (Thelypteridaceae) have certain natural distribution in the subtropical areas of the country, but are more frequently found as escaped plants from greenhouses. Murakami et al. (2008) demonstrated a range expansion of Thelypteris dentata facilitated by urbanization in Kinki District, Japan. The natural habitat of Cyrtomium falcatum subsp. falcatum is restricted to coastal areas (Ebihara et al. 2017a, Matsumoto 2003), and scattered occurrences in urban environments (Ebihara 2017) in inlands are probably introduced ones.

## Reticulate evolution

Species complexes or aggregate species exhibiting wide ranges of continuous morphological variation are quite commonly known in ferns and lycophytes (Haufler 1996), and this trend seems to be true for the flora of Japan. In case of that interspecific F1 hybrids make the species boundaries ambiguous, we could usually find distinct gaps between them by removing their hybrid individuals suggested by irregular-shaped spores with low germination rates (e.g. Ohta and Takamiya 1999 in the Diplazium mettenianum complex). When interspecific hybrids overcome sterility by chromosome doubling, the situation becomes more complicated and the biological units composing the complex are not easily clarifiable solely by morphological approach. The allopolyploids of hybrid origins often re-hybridize with their parental species --- such "reticulate evolution" process became well-known after the study on Appalachain Asplenium (Wagner 1954). Apparently similar approach combining chromosome and morphological observations was rarely adopted to Japanese species. After 1990's, allozyme analyses were frequently applied to the fern complexes in Japan, and plastid sequences often accompanied as indicators of maternal lineage (e.g. Darnaedi et al. 1990 in Dryopteris yakusilvicola and its related species, Lin et al. 1996 in Odontosoria, Shinohara et al. 2010 in Lepisorus, Takamiya and Ohta 2001 in Diplazium). In 2000's, single or low copy nuclear DNA regions became popularly used as biparently inherited markers (e.g. Ishikawa et al. 2002), and sequence polymorphisms were efficiently isolated by cloning or the SSCP (Single Strand Conformation Polymorphism) method (initiated by Ebihara et al. 2005 in the Vandenboschia radicans complex, thereafter applied to various fern groups including Ceratopteris [Adjie et al. 2007]), the Pteris cretica complex [Jaruwattanaphan et al. 2013], the Davallia repens complex [Chen et al. 2014], the Asplenium normale complex [Fujiwara et al. 2017]). Although next-generation sequencing is widely used in phylogenetic and population genetic studies on various organisms in recent years, it is still not popularly applied to analyses on species complexes of ferns and lycophytes mainly due to technical difficulties in removing chimeral contigs identifying and homologous copies from the obtained sequences. More recently Rothfels et al. (2017) succeeded in phylogenetic analysis on a polyploid complex of cystoptreroid ferns using PacBio sequences, and the method is highly expected to apply wide ranges of fern complexes.

## Species complexes involved with apomixis

Further complexity is provided when apomictic reproduction (reviewed in Grusz et al. 2016) is concerned. The topic, apomixis or apogamy in ferns has attracted researchers in Japan as the reproductive mode is known in a higher rate in the native taxa of the country (13 % of the species whose reproductive modes are known in Takamiya 1996; 12 % in present study) than the average rate in the world (ca. 10 %, Grusz et al. 2016). Apomixis plays a role not only in overcoming sterility of interspecific hybrids (e.g. Park and Kato 2003) but also making secondary hybridization by fertilization between a sperm produced by the ‘hybrid’ and an egg of sexually reproducing races. This process can generates countless apomict races with various genomic combinations which are morphologically continuous, and the races are often controversial about circumscription of ‘species’ (a polymorphic species versus numerous morphologically ill-defined species). One of the best known apomict complexes troublesome in identification, the Dryopteris varia complex was recently clarified their internal relationships using single-copy nuclear DNA markers (Hori et al. 2014, 2015, 2016), and analyses on several other complexes are in progress. However, even using molecular markers, we often meet an obstacle that we cannot find progenitors or sexual diploids of the complex. Hori et al. (2018d) discovered several sexual diploid individuals of Dryopteris in southern China, and as the progenitors the newly found diploids well explained the reticulate evolution processes of the D. erythrosora complex in Japan which had bothered researchers for many years (Hori et al. 2018b). This case supported the hypothesis by Gao et al (2003): “The ancestral sexual species of many Japanese apogamous species occur in China, suggesting that these species might have migrated from China”. In contrast to these successful clarifications, we are still struggling with more challenging complexes. The species of Cyrtomium (Dryopteridaceae) in Japan typically match this pattern: 12 native species and subspecies are all triploid apomicts except for two subspecies of C. falcatum (subsp. australe and subsp. littorale, sexual diploid) and C. devexiscapulae (sexual tetraploid) (Ebihara et al. 2017a, Matsumoto 2003). Despite careful exploration of sexually reproducing individuals by examining spore number per sporangium in herbarium specimens collected in China, a large part of them are apomictic and sexual ones were found less frequently in geographically narrow ranges (Mitsuta 1986, 1988). The Pteris fauriei complex is probably in similar situation, and it includes several endemic ‘species’ to Japan which are diploid or triploid apomict (Ebihara 2016). The “Diploids-First Approach” (Beck et al. 2010) is a principle for the studies on fern reticulate complexes, but we should mind that diploids are not always present nearby including possible extinction in complexes involved with apomixis.

## Species concept
